### 2021-03-20 paper.Rmd
compress intro paragraph and reorder

@@

While tours are invaluable for assessing the geometry of data, they are by no means the only technique available for visualising structure in high dimensional data. An early technique proposed for assessing pairwise relationships between variables is the scatter plot matrix (SPLOM) [@Tukey1983-fj] <!-- there might be others here(?) -->. In this graphic, the canvas is split into a matrix where each off-diagonal element is a scatter plot of variable pairs, while the diagonal elements are either left blank or represent univariate structure. The SPLOM allows the viewer to assess correlation structure but does not scale to large numbers of variables.

Similarly, parallel coordinates plots (PCP) can be used to explore correlation and co-linearities in multivariate data [@Inselberg1985-kf]. In PCPs, variables are arranged along the (vertical) horizontal axis with vertical (horizontal) axes drawn representing the range of values for each variable. Points (observations) of the data are represented by drawing lines between each variables axis. If there is positive correlation between two variables the lines will be parallel, while for negative correlation the lines will cross. By placing multiple variables side by side in a PCP higher order structure like clustering or lower dimensional embeddings is revealed, however the ordering of variables is important to reveal these features.

The heatmap display is widely used to visualise cluster structure in bioinformatics. In a heatmap each element of a numeric matrix is displayed as a single colored square pixel on the canvas. The hue of the color is mapped to the magnitude of the element's values. A subtlety of using the heatmap is determining how to order the rows and columns of the matrix called seriation. In practice, two way hierarchical clustering  or singular value decomposition is used to order rows and columns. See @Wilkinson2009-pj for a comprehensive history.

<!-- above paragraphs could be shortened into a single paragraph? -->

<!-- 
Not sure how much of this wee need to go into...
- subplots trellis literature
- data pipelines (?)
- What do we mean by multivariate data visualisation?
- multivariate categorical vis, tree maps etc.
-->

>>>>

While tours are invaluable for assessing the geometry of data, they are by no means the only technique available for visualising structure in high dimensional data. An early technique proposed for assessing pairwise relationships between variables is the scatter plot matrix (SPLOM) [@Tukey1981-ov; @Chambers1983-dh; @Tukey1983-fj; @Carr1984-xw; @Becker1987-dw]. The SPLOM allows the viewer to assess correlation structure but does not scale to large numbers of variables. Similarly, parallel coordinates plots (PCP) can be used to explore correlation and multicollinearity [@Inselberg1985-kf; @Wegman1990-dy]. By placing multiple variables side by side in a PCP higher order structure like clustering or lower dimensional embeddings is revealed, however the ordering of variables along the axis changes what can be learned. Another display that relies on variable ordering is the heatmap, which is widely used to visualise cluster structure in bioinformatics. @Wilkinson2009-pj provides a comprehensive history of this display in the social and natural sciences.

@@
